---
title: "Factors Influencing Student JAMB Scores"
author: "Group Members: Alicia, Bruce, Kyle, Victor"
date: "2024/11/09"
output: 
  beamer_presentation:
    slide_level: 2
---

 
The primary objective of this study is to investigate the factors that influence students' JAMB scores (Joint Admissions and Matriculation Board scores). Specifically, we aim to answer the following questions:

- How do study habits and attendance affect JAMB scores?
- What is the impact of socioeconomic factors on students' academic performance in the JAMB examination?
- Are there significant differences in JAMB scores based on school type or location?

# Introduction

## Background

The JAMB examination is a standardized test in Nigeria used for admission into tertiary institutions. Students' performance in this exam is crucial for their academic progression. Various factors such as study habits, school environment, and socioeconomic status may influence JAMB scores. Prior research suggests that factors like study hours, attendance, and parental involvement significantly impact standardized test scores.



## Analysis Plan

We plan to use linear regression models to analyze the relationship between the predictors and the JAMB scores. We will:

- Examine the data with visualization
- Use forward stepwise selection to identify significant predictors.
- Validate the final model using diagnostic tests.

# Data Collection

## Dataset Description

- **Source:** Collected from a survey of high school students preparing for the JAMB examination.
- **Sample Size:** 5000 students.
- **Variables:**

  - **Dependent Variable:** `JAMB_Score` (Students' scores in the JAMB examination, 100-400).
  - **Independent Variables:**
    - `Study_Hours_Per_Week`
    - `Attendance_Rate`
    - `Teacher_Quality` (Scale 1-5)
    - `Distance_To_School`
    - `School_Type` (Public/Private)
    - `School_Location` (Urban/Rural)
    - `Extra_Tutorials` (Yes/No)
    - `Access_To_Learning_Materials` (Yes/No)
    - `Parent_Involvement` (Low/Medium/High)
    - `IT_Knowledge` (Low/Medium/High)
    - `Age`
    - `Gender` (Male/Female)
    - `Socioeconomic_Status` (Low/Medium/High)
    - `Parent_Education_Level` (Primary/Secondary/Tertiary)

### Loading the Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages
library(car)
library(leaps)
library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)
library(ggcorrplot)
library(GGally)

# Read the data
data <- read.csv("jamb_exam_results.csv")
```

\section{Data Exploration}

## Distribution

```{r, echo=FALSE}
## Distribution of JAMB Scores
data <- read.csv("jamb_exam_results.csv")
ggplot(data, aes(x = JAMB_Score)) +
  geom_histogram(binwidth = 10, fill = "skyblue", color = "black") +
  labs(title = "Distribution of JAMB Scores", x = "JAMB Score", y = "Count")

```

## Correlation Matrix

```{r, echo=FALSE}
## Extended Correlation Matrix with Additional Factors

library(ggcorrplot)

numerical_vars <- data %>%
  select(JAMB_Score, Study_Hours_Per_Week, Attendance_Rate, Teacher_Quality, 
         Distance_To_School, Age, Assignments_Completed, Teacher_Quality, Attendance_Rate)

corr <- round(cor(numerical_vars), 1)

ggcorrplot(
  corr,
  lab = TRUE,
  title = "Correlation Matrix"
)


```
## Gender
```{r, echo=FALSE}
## Distribution of JAMB Scores Separated by Gender (Density Plot)

ggplot(data, aes(x = JAMB_Score, fill = Gender)) +
  geom_density(alpha = 0.7) +  
  scale_fill_manual(values = c("lightblue", "lightpink")) +
  labs(
    title = "Distribution of JAMB Scores by Gender",
    x = "JAMB Score",
    y = "Density",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

```


## Socioeconomic Status


```{r, echo=FALSE}
## JAMB Score by Socioeconomic Status and Parent Education Level

data$Socioeconomic_Status <- factor(data$Socioeconomic_Status, levels = c("Low", "Medium", "High"))

ggplot(data, aes(x = Socioeconomic_Status, y = JAMB_Score, fill = Parent_Education_Level)) +
  geom_boxplot() +
  labs(title = "JAMB Score by Socioeconomic Status and Parent Education Level", 
       x = "Socioeconomic Status", 
       y = "JAMB Score")


```

## School location and type

```{r, echo=FALSE}
## Distribution of JAMB Scores by School Location and School Type

library(ggplot2)

ggplot(data, aes(x = School_Location, y = JAMB_Score, fill = School_Type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.6) + 
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red", 
               position = position_dodge(width = 0.6)) + 
  labs(
    title = "JAMB Scores by School Location and School Type",
    x = "School Location",
    y = "JAMB Score",
    fill = "School Type"
  ) +
  scale_fill_brewer(palette = "Dark2") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14), 
    legend.position = "top",  
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )


```

##  Teacher Quality
```{r, echo=FALSE}


library(ggplot2)

ggplot(data, aes(x = factor(Teacher_Quality), y = JAMB_Score, fill = factor(Teacher_Quality))) +
  geom_violin(trim = FALSE, alpha = 0.7) +  
  geom_boxplot(width = 0.2, outlier.shape = NA, color = "black", alpha = 0.5) +
  scale_fill_brewer(palette = "Reds") + 
  labs(
    title = "JAMB Scores by  Teacher Quality",
    x = "Teacher Quality (1 = Low, 5 = High)",
    y = "JAMB Score",
    fill = "Teacher Quality"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14),  
    legend.position = "right" 
  )

```

## Study Hours and Attendance Rate
```{r, echo=FALSE}


library(ggplot2)

ggplot(data, aes(x = Study_Hours_Per_Week, y = Attendance_Rate, size = JAMB_Score, color = JAMB_Score)) +
  geom_point(alpha = 0.6) +  
  scale_color_gradient(low = "blue", high = "red") +  
  labs(
    title = "JAMB Scores by Study Hours and Attendance Rate",
    x = "Study Hours Per Week",
    y = "Attendance Rate (%)",
    size = "JAMB Score",
    color = "JAMB Score"
  ) +
  theme_minimal() +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.text = element_text(size = 12),  
    axis.title = element_text(size = 14)  
  )

```

## Extra Tutorials and Access to Learning Materials
```{r, echo=FALSE}
## Extra Tutorials and Access to Learning Materials

data$Tutorials_Materials <- paste(data$Extra_Tutorials, data$Access_To_Learning_Materials, sep = ", ")

mean_scores <- data %>%
  group_by(Tutorials_Materials) %>%
  summarise(Mean_Score = mean(JAMB_Score))

library(ggplot2)

ggplot(data, aes(x = Tutorials_Materials, y = JAMB_Score, fill = Tutorials_Materials)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +  
  geom_point(data = mean_scores, aes(x = Tutorials_Materials, y = Mean_Score), 
             color = "#EE5C42", size = 3, shape = 18) + 
  geom_text(data = mean_scores, aes(x = Tutorials_Materials, y = Mean_Score, label = round(Mean_Score, 1)), 
            vjust = -0.5, color = "#EE5C42", size = 4) +  
  scale_fill_brewer(palette = "GnBu") +
  labs(
    title = "Extra Tutorials and Access to Learning Materials",
    x = "Combination of Extra Tutorials and Access to Learning Materials",
    y = "JAMB Score",
    fill = "Combination"
  ) +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14) 
  )
data$Tutorials_Materials <- NULL
```

## Distance to School
```{r, echo=FALSE}
## Distance to School

library(ggplot2)

ggplot(data, aes(x = Distance_To_School, y = JAMB_Score)) +
  geom_point(aes(color = Distance_To_School), alpha = 0.6, size = 2) + 
  geom_smooth(method = "lm", se = TRUE, color = "#97FFFF", linetype = "dashed") +
  scale_color_gradient(low = "green", high = "red") +
  labs(
    title = "JAMB Scores by Distance to School",
    x = "Distance to School (km)",
    y = "JAMB Score",
    color = "Distance"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.text = element_text(size = 12), 
    axis.title = element_text(size = 14) 
  )

```


##  Ages
```{r, echo=FALSE}
##  Ages

library(ggplot2)

ggplot(data, aes(x = factor(Age), y = JAMB_Score, fill = factor(Age))) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) + 
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +  
  scale_fill_viridis_d(option = "plasma") + 
  labs(
    title = "JAMB Score Distribution by Age",
    x = "Age",
    y = "JAMB Score",
    fill = "Age"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.text = element_text(size = 12),  
    axis.title = element_text(size = 14),  
    legend.position = "none" 
  )

```


## Parental Involvement
```{r, echo=FALSE}
##  Involvement
data$Parent_Involvement <- factor(data$Parent_Involvement, levels = c("Low", "Medium", "High"))

library(ggplot2)

ggplot(data, aes(x = Parent_Involvement, y = JAMB_Score, fill = Parent_Involvement)) +
  geom_violin(trim = FALSE, alpha = 0.7) + 
  geom_boxplot(width = 0.2, color = "black", alpha = 0.5, outlier.shape = NA) +  
  scale_fill_brewer(palette = "Accent") +
  labs(
    title = "JAMB Scores by Parent Involvement",
    x = "Parent Involvement Level",
    y = "JAMB Score",
    fill = "Parent Involvement"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none" 
  )

```


## IT Knowledge
```{r, echo=FALSE}
## IT Knowledge


data$IT_Knowledge <- factor(data$IT_Knowledge, levels = c("Low", "Medium", "High"))

library(ggplot2)

ggplot(data, aes(x = IT_Knowledge, y = JAMB_Score, fill = IT_Knowledge)) +
  geom_violin(trim = FALSE, alpha = 0.7) +  
  geom_boxplot(width = 0.2, color = "black", alpha = 0.5, outlier.shape = NA) +  # Boxplot overlay
  scale_fill_brewer(palette = "Set2") + 
  labs(
    title = "JAMB Scores by IT Knowledge",
    x = "IT Knowledge Level",
    y = "JAMB Score",
    fill = "IT Knowledge"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "none" 
  )

```



# Data Analysis and Results
Before fitting any model, we must first preprocess the data. There are a few things we need to handle:

- A lot of the features are categorical, so we need to transform them into numerical data using R's `factor` function.
- JAMB scores have a range from 100 to 400 (and scores above 300 is very rare,) so near the bottom and the top quintile of the data, the residuals might be capped and we'd have issue with Homoscedasticity. Indeed, when we fit an OLS model with the whole data set, we see the Q-Q Residual graph suggesting thinner tails than normal. This is alleviated by discarding the top and bottom 10% of the data.



```{r, echo=FALSE}
# Convert categorical variables to dummy variables
# Convert columns to factors with specified levels and orders
data$School_Type <- factor(data$School_Type)
data$School_Location <- factor(data$School_Location)
data$Extra_Tutorials <- factor(data$Extra_Tutorials)
data$Access_To_Learning_Materials <- factor(data$Access_To_Learning_Materials)

# Ordered factors with specified levels
data$Parent_Involvement <- factor(data$Parent_Involvement, levels = c("Low", "Medium", "High"), ordered = TRUE)
data$IT_Knowledge <- factor(data$IT_Knowledge, levels = c("Low", "Medium", "High"), ordered = TRUE)
data$Gender <- factor(data$Gender)
data$Socioeconomic_Status <- factor(data$Socioeconomic_Status, levels = c("Low", "Medium", "High"), ordered = TRUE)
data$Parent_Education_Level <- factor(data$Parent_Education_Level, levels = c("Primary", "Secondary", "Tertiary"), ordered = TRUE)

# transforming jamb scores into percentiles

# data$JAMB_Score_Percentile <- rank(data$JAMB_Score) / 5000 * 100 

# Remove outliers from the 5th to 95th percentile for continuous variables
remove_outliers <- function(data, column) {
  lower_bound <- quantile(data[[column]], 0.10, na.rm = TRUE)
  upper_bound <- quantile(data[[column]], 0.90, na.rm = TRUE)
  data <- data[data[[column]] >= lower_bound & data[[column]] <= upper_bound, ]
  return(data)
}

# Apply the function to continuous variables (e.g., JAMB_Score)
clean_data <- data
clean_data <- remove_outliers(clean_data, "JAMB_Score")

# Continue with the cleaned data
clean_data <- na.omit(clean_data)


```

## Initial Model
\tiny
```{r, echo=FALSE}
# initial model with all predictors, ordinary linear regression
initial_model <- lm(JAMB_Score ~ ., data = clean_data)
summary(initial_model)

```
\normalsize

# Model Selection
Here we utilize a stepwise selection method based on AIC. We start with the initial model with all the features, and each step we add or subtract one predictor at a time that minimizes the Akaike Information Criterion. We repeat this until there's no improvement possible.

### Stepwise Selection
```{r, include=FALSE}
library(MASS)
```

```{r, include=FALSE}
# Using backward stepwise selection based on AIC


step_model <- stepAIC(initial_model)
```

# Final Model Diagnosics

## Residual Analysis
We can see here that the Q-Q plot still suggests a thin tail. The effect of a capped target is also noticable in the Residual vs Fitted Graph.
```{r, echo=FALSE}
# Plot residuals
par(mfrow = c(2, 2))
plot(step_model)
```


## Normality of Residuals
Here we test the normality of residuals. We have little evidence that the true distribution of test performances are Gaussian, so we aren't surprised by the results.
```{r, echo=FALSE}
shapiro.test(residuals(step_model))
```

## Homoscedasticity
The test fails to reject the null hypothesis that the data is homoscedastic. This is in part due to us clipping the data.
```{r, include=FALSE}
# Breusch-Pagan test
library(lmtest)
```

```{r, echo=FALSE}
bptest(step_model)
```

## Multicollinearity, Final model
None of the included features in the final model are heavily collinear. Of course this is due to us using stepwise search and excluding collinear features.
\small
```{r, echo=FALSE}
# Variance Inflation Factor (VIF)
vif(step_model)
```
\normalsize
## Multicollinearity, initial model
Other than a good amount of correlation between study hours per week and assignment completion rate, the features are relatively orthogonal.
\small
```{r, echo=FALSE}
# Variance Inflation Factor (VIF)
vif(initial_model)
```
\normalsize

# Final Regression Model

## Model Summary
\tiny
```{r, echo=FALSE}
# Capture the summary output as a character vector
summary_text <- capture.output(summary(step_model))

# Display the summary, starting from the 11th line
cat(summary_text[11:length(summary_text)], sep = "\n")
```
\normalsize

## Confidence Intervals
\tiny
```{r}
# Calculate confidence intervals for the final model
final_conf_intervals <- confint(step_model)

# Print the confidence intervals
print(final_conf_intervals)
```
\normalsize

## Conclusion




